Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2021-05-03 paper/response/AJPRefereeResponse.tex

We thank you and the referees for your helpful comments and suggestions. 
We address the comments below in turn.

\subsection*{Comments from editor, Joseph Amato}
\begin{quote}
This is an addendum to my decision letter and the two reviews of your manuscript.  I am satisfied that the reviewers read your work carefully, and that their recommendations are fair, thoughtful, and potentially useful.  While I sympathize with R1’s concern that the manuscript is better suited to a research journal, I think it could be revised to make a valuable contribution to AJP, which, as you know, is geared toward pedagogy.  So as I read your manuscript, I recorded things that were not sufficiently clear to me, supposing that the same things might not be clear to many AJP readers who might be very interested in duplicating your apparatus.  Here are some of my concerns.
\end{quote}

\noindent 1. Simple things: Convert footnotes to endnotes, blending them in with the bibliography.  Also, use US English (color, not colour; defense, not defence, etc.). 
\begin{itemize}
\item We have moved the footnotes to the bibliography
\item \todo{Work in progress, please correct any that have not been fixed yet}
\end{itemize}

\noindent 2. Section III.  How was the Fourier transform generated?  (software?)
\begin{itemize}
\item Yes, the Fourier transform was generated using software. We have added a sentence to Section III (``The Discrete Fourier transform was calculated using the NumPy package for Python.''), a citation (to the creators of NumPy), and a footnote (to the documentation).
\end{itemize}

\noindent 3. Section IV.  The term “Markov process” may not be familiar to readers.  Figure 4: The three shades (or are there four?) indicating probability do not show up well in color, and worse, will be indistinguishable in black/white.  I’m guessing that’s the reason I didn’t understand why certain paths were grey and not black.  While I applaud your efforts to explain the Viterbi algorithm (VA), I still think there’s room for improvement.  I don’t understand how it improves the signal to noise ratio. 
\begin{itemize}
\item We have added a sentence to explain Markov process: ``In a Markov process, the current state depends only on the previous state (in this case the state is the frequency of the signal).''
\item We have lightened the colour scheme used in the Fig. 4 schematic to aid with black and white printing. Our intention is that the Fig. 4 schematic be interpreted as a simplified version of the Fig. 5 spectrogram result. To clarify this we have made the colour scheme of the schematic more consistent with that of Fig. 5 and we have refered back to the schematic when we first introduce Fig. 5. 
\item The hidden Markov model and Viterbi algorithm's strength is in it's ability to find randomly wandering signals. The single constant tone in section III is recovered easily using a Fourier transform, however if the signal frequency changes over time, the signal power is spread across multiple frequency bins. We have added a new figure to illustrate this (Fig. 4) and some text at the start of Section IV (paragraph 2) and at the end of Section IV. B. to explain that the Fourier transform of the whole dataset is not well suited to this type of signal.
\item We have also revised the entire Section IV to improve clarity in our Viterbi algorithm explanation and motivation.
\end{itemize}

\noindent 4. Figure 5 is described as a heatmap.  Please explain, in the caption, what the color coding indicates.  Can you compare what is shown to what would be detected without using the VA?  Is this ``heatmap'' the same representation as is often shown in LIGO publications?
\begin{itemize}
\item The color coding represents the logarithm of the absolute value of the discrete Fourier transform. We have clarified this in the caption.
\item Without using the VA the signal becomes spread out over the frequency bins and might not rise above the noise. We have now explained this and added a plot that shows the Fourier spectrum of the entire timeseries.
\item The heatmap is similar to the figures in LIGO-Virgo-KAGRA publications which show transient gravitational-wave signals, but not identical. Those figures normally show Q-transforms [described in Chatterji, et al., Classical Quantum Gravity 21, S1809 (2004)] whereas we show the Fourier transform in this work. We have not drawn comparisons the the Q-transform plots in order to avoid confusion.
\end{itemize}

\noindent 5. Section V.  What do you mean by an optical microphone device?  “Cloth screen” over the photodiode?  Fig. 6 is unnecessary; the differences from Fig. 1 are explained clearly in the text.
\begin{itemize}
\item Optical microphone: we have rearranged the text so that the optical microphone concept is introduced in a separate paragraph and expanded our discission of it to improve clarity. (Section V, paragraph 2)
\item Fig. 6: agreed, we have removed this figure and moved the photo to the Supplementary Material.
\item The cloth screen was used to reduce the intensity of light on the photodiode and avoid saturating the analog-to-digital converter. It was indeed an improvised solution using the ``grill cloth'' and plastic frame from the dismantelled commercial speaker used in the demonstration. We have added a sentence to Section V.A. to clarify.
\end{itemize}

@@

\noindent
-- Is the HMM defined properly in Sec. IV A? The observables here are the amplitudes of the spectrogram pixels at a given time index and the hidden markov process is the underlying drift of frequency and amplitude of the signal. I was expecting to see a conditional pdf that connects the state of the underlying (suitably discretized) process to the observables. I don't see it.
\begin{itemize}
\item We have clarified the explanation of the HMM in the text. The detection statistic captures the likelihood of a signal being present in each time-frequency bin given the observable data. In Section IV.A. we have expanded our description of gravitational wave analysis and referred to both the transmission matrix and emission matrix by name to reduce confusion. 
\end{itemize}

\noindent
-- Is Eq. 1 correct? The LHS is a marginal probability while the RHS is a joint probability. Shouldn't there be a marginalization on the RHS?
\begin{itemize}
\item This equation was indeed incorrect and we thank you for the careful review. The equation should give the probability for a specific path, and is not a sum over all possibilities. We have corrected Eq. 1 and added clarification to the text. 
\end{itemize}

\noindent
-- The normalization used for $F(t_i, f_j)$ should be defined explicitly. 
\begin{itemize}
\item Each value is divided by the maximum value in the grid such that $\max_{i,j} F(t_i, f_j) = 1$. We have added this information to section IV.B.
\end{itemize}

### 2021-05-03 paper/ifo-singleTone.tex
fixing plots - part 1
fixing and combining plots - part 2
defined normalisation, added comparison spectrum of entire timeseries for VA discussion, grammar/spelling check
working on single and wandering tone text and response

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{\label{fig:webcam_spectrum}
Recovery of a tone at a constant frequency: Here we plot the Fourier amplitude of the intensity pattern against frequency.
The injected signal has a frequency of $2.09\,{\rm Hz}$, while the recovered signal peaks at $2.099\,{\rm Hz}$ and has an FWHM of $0.033\,{\rm Hz}$.
The plot also shows two harmonics at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ with amplitudes $10.3 \%$ and $9.8 \%$ of the amplitude of the primary peak, respectively.
}	
\end{figure}

% rearranged as I found this section to switch between method and results too much, I've tried this order now: intro, webcam and data specifics, injection, and results.

@@

As described in Section~\ref{sec:ifo}, the audio signal is played through a speaker fixed to the back of M2. 
The intensity of the interference pattern is measured at a single point on the screen, indicated by the pink star in Fig~\ref{fig:interference_pattern}. 
The webcam records video in three color channels; red, blue, and green. 
We use the green channel as an approximation of the total intensity produced by the monochromatic green laser.
The webcam samples at a rate of $30\, {\rm Hz}$, which limits the spectral content of observable signals to less than $15\,{\rm Hz}$, the Nyquist frequency.

@@

A tone with a frequency of $2.09\,{\rm Hz}$ is played through the speaker for one minute. 
The amplitude of the Fourier transform of the recovered signal is shown in Fig.~\ref{fig:webcam_spectrum}. The Discrete Fourier transform was calculated using the NumPy package for Python~\cite{numpy}~\footnote{\url{https://numpy.org/doc/stable/reference/routines.fft.html}}.
We measured a peak amplitude at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$.
Two harmonics can also be seen at integer multiples of the peak frequency. 
The peaks at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ have amplitudes of $0.103$ and $0.098$ as a fraction of the height of the main peak, respectively. They are likely due to the nonlinear response of the system discussed in Section~\ref{sec:ifo} and Appendix~\ref{app:intensity_derivation}. Also note that the noise appears to rise at lower frequencies.

In the following section, we use a hidden Markov model technique, commonly used in continuous-wave searches, to recover a slowly wandering frequency signal.

>>>>

\begin{figure}
	\includegraphics[width=.45\textwidth]{figures/webcam_spectrum_expt_4_0209.pdf}
	\caption{\label{fig:webcam_spectrum}
Recovery of a tone at a constant frequency: Here we plot the Fourier amplitude of the intensity pattern against frequency.
The injected signal has a frequency of $2.09\,{\rm Hz}$, while the recovered signal peaks at $2.099\,{\rm Hz}$ and has an FWHM of $0.033\,{\rm Hz}$.
The plot also shows two harmonics at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ with amplitudes $10.3 \%$ and $9.8 \%$ of the amplitude of the primary peak, respectively.
}	
\end{figure}

% rearranged as I found this section to not switch between method and results too much, I've tried this order now: intro, webcam and data specifics, injection, and results.

@@

As described in Section~\ref{sec:ifo}, the audio signal is played through a speaker fixed to the back of M2 (see Fig.~\ref{fig:ifo_schematic_webcam}). 
The intensity of the interference pattern is measured at a single point on the screen, indicated by the pink star in Fig.~\ref{fig:interference_pattern}. 
The webcam records video in three color channels; red, blue, and green. 
We use the green channel as an approximation of the total intensity produced by the monochromatic green laser.
The webcam samples at a rate of $30\, {\rm Hz}$, which limits the spectral content of observable signals to less than $15\,{\rm Hz}$, the Nyquist frequency.

@@

A tone with a frequency of $2.09\,{\rm Hz}$ is played through the speaker for one minute. 
The amplitude of the Fourier transform of the recovered signal is shown in Fig.~\ref{fig:webcam_spectrum}. The Discrete Fourier transform was calculated using the NumPy package for Python~\cite{numpy}~\footnote{\url{https://numpy.org/doc/stable/reference/routines.fft.html}}.
We measured a peak amplitude at $2.099\,{\rm Hz}$ with a full width half maximum (FWHM) of $0.033\,{\rm Hz}$.
Two harmonics can also be seen at integer multiples of the peak frequency. 
The peaks at $4.19\,{\rm Hz}$ and $6.28\,{\rm Hz}$ have amplitudes of $0.103$ and $0.098$ as a fraction of the height of the main peak, respectively. They are likely due to the nonlinear response of the system discussed in Section~\ref{sec:ifo} and Appendix~\ref{app:intensity_derivation}. Also, note that the noise appears to rise at lower frequencies.

In the following section, we use a hidden Markov model technique, commonly used in continuous-wave searches, to recover a signal with a slowly changing frequency.

### 2021-05-03 paper/ifo-wanderingTone.tex
working on single and wandering tone text and response

\begin{figure}
	\includegraphics[width=.45\textwidth]{figures/viterbi_comparison_entire_spectrum_viterbi_test_webcam.pdf}
	\caption{\label{fig:viteri_comparison_webcam_spectrum}
Why we use the Viterbi algorithm to recover wandering frequency signals: Here we plot the Fourier amplitude of the intensity pattern against frequency as in Fig.~\ref{fig:webcam_spectrum} for the entire timeseries of the wandering frequency signal in Fig.~\ref{fig:viterbi_overlay}. Instead of resolving definite peaks like the constant frequency signal shown in Fig.~\ref{fig:webcam_spectrum}, we only see inconclusive peaks at 5.7~Hz and 10.7~Hz that correspond to the long amount of time that the wandering frequency spends near those values and/or its amplitude there, as seen in the pink dots in Fig.~\ref{fig:viterbi_overlay}. With the signal spread out over the frequency bins, we are unable to reliably detect it above the noise and therefore we use the Viterbi algorithm instead.
The large values at low frequencies below 0.1~Hz are similar to those in Fig.~\ref{fig:webcam_spectrum} and can be ignored as they are likely inaccurate due to the short length of the signal relative to their period (i.e.\ 300~s only contains 30~periods of a 0.1~Hz signal), but this feature would require further research to fully determine its origin. %maybe could be from the long period (100~s) variation?
}
\end{figure}

@@

A Fourier transform applied to the whole dataset (as in Section~\ref{sec:single_tone}) is not well suited to the wandering frequency case, as the signal can be spread across multiple frequency bins making it challenging or perhaps impossible to detect, as shown in Fig.~\ref{fig:viteri_comparison_webcam_spectrum}.
%A different means of detection is needed for a signal which has a randomly wandering frequency.

In continuous wave searches, the wandering frequency problem is solved using the Viterbi algorithm~\cite{Viterbi:1967}, which can track the signal's frequency over time~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}. 
The analysis described and presented in this section is directly inspired by real continuous-wave searches~\cite{} 
In section~\ref{sec:realCWSearches} we briefly review the analysis method used by LIGO and Virgo, in section~\ref{} we describe the methods as applied in this work, and in section~\cite{} we our results of recovering a wandering frequency signal using the table-top interferometer.

@@

The methods used here are directly inspired by LIGO-Virgo continuous wave searches. 
For further details on these methods see Refs.~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017} and for searches performed using them see Refs.\cite{ScoX1O2Viterbi:2019, ScoX1ViterbiO1:2017, MillhouseStrangMelatos:2020, JonesSun:2020, MiddletonEtAlO2LMXBs:2020, PostMergerRemnantSearch:2019, SunEtAlSNR:2018, viterbi_application}. \han{check for new papers}

@@

In continuous wave searches, the signal amplitude is expected to be small in comparison to the noise and its frequency could be changing unpredictably over time.
A Fourier transform of the entire dataset as described in Section~\ref{sec:single_tone} would split the signal power between multiple frequency bins and could result in the signal being undetectable under the noise. 
The Viterbi algorithm's strength lies in its ability to track such signals through the data.
In the following section, we present the results of using the Viterbi algorithm with the table-top interferometer data. 
% - just simiplifed this discussion a little and added a reference to the CW searches
%The key to the Viterbi algorithm is that we can use the signal's correlation in time to improve the signal-to-noise ratio of the frequency of the wandering signal at a particular time. Since the noise is assumed to not be correlated in time, if the detection statistic at a given frequency is high at a particular time and at neighboring frequencies at neighboring times then the likelihood that the given frequency at that particular time is from the wandering signal increases. By iterating this process through the grid for a suitable choice of the window of neighboring frequencies and times, the Viterbi algorithm is well-suited to finding wandering signals.

>>>>

@@

We first demonstrate why a Fourier transform applied to the whole dataset (as in Section~\ref{sec:single_tone}) is not well suited to the wandering frequency case.
As shown in Fig.~\ref{fig:viterbi_comparison_webcam_spectrum}, the signal is spread across multiple frequency bins, making it challenging or in some cases impossible to detect.
%A different means of detection is needed for a signal which has a randomly wandering frequency.

In continuous wave searches, the wandering frequency problem is solved using the Viterbi algorithm~\cite{Viterbi:1967}, which can track the signal's frequency over time.
The analysis described and presented in this section is directly inspired by real continuous-wave search methods~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}. 
In section~\ref{sec:realCWSearches} we briefly review the analysis method used by LIGO and Virgo, in Section~\ref{sec:viterbi} we describe the methods as applied in this work, and in section~\ref{sec:wanderingResults} we show results for recovering a wandering frequency signal using the table-top interferometer.

\begin{figure}
	\includegraphics[width=.45\textwidth]{figures/viterbi_comparison_entire_spectrum_viterbi_test_webcam.pdf}
	\caption{\label{fig:viterbi_comparison_webcam_spectrum}
Fourier amplitude identical to that shown in Fig.~\ref{fig:webcam_spectrum}, except using a signal which has a slowly changing frequency instead of a constant frequency signal.
The signal is spread between many frequency bins. 
We see peaks at $5.7\,{\rm Hz}$ and $10.7\,{\rm Hz}$ and refer the reader forward to Fig.~\ref{fig:viterbi_overlay} to see the frequency evolution of this signal. 
A single Fourier transform of the entire dataset is not well suited to detecting signals with wandering frequencies.
% just making this a little more concise 
%Why we use the Viterbi algorithm to recover wandering frequency signals: Here we plot the Fourier amplitude of the intensity pattern against frequency as in Fig.~\ref{fig:webcam_spectrum} for the entire timeseries of the wandering frequency signal in Fig.~\ref{fig:viterbi_overlay}. Instead of resolving definite peaks like the constant frequency signal shown in Fig.~\ref{fig:webcam_spectrum}, we only see inconclusive peaks at 5.7~Hz and 10.7~Hz that correspond to the long amount of time that the wandering frequency spends near those values and/or its amplitude there, as seen in the pink dots in Fig.~\ref{fig:viterbi_overlay}. With the signal spread out over the frequency bins, we are unable to reliably detect it above the noise and therefore we use the Viterbi algorithm instead.
%The large values at low frequencies below 0.1~Hz are similar to those in Fig.~\ref{fig:webcam_spectrum} and can be ignored as they are likely inaccurate due to the short length of the signal relative to their period (i.e.\ 300~s only contains 30~periods of a 0.1~Hz signal), but this feature would require further research to fully determine its origin. %maybe could be from the long period (100~s) variation?
}
\end{figure}

@@

The methods used here are directly inspired by LIGO-Virgo continuous wave searches. 
For further details on these methods see Refs.~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017} and for searches performed using them see Refs.\cite{ScoX1O2Viterbi:2019, ScoX1ViterbiO1:2017, MillhouseStrangMelatos:2020, JonesSun:2020, MiddletonEtAlO2LMXBs:2020, PostMergerRemnantSearch:2019, SunEtAlSNR:2018, viterbi_application}. \han{check for new papers - add Deeksha's paper}

@@

In continuous wave searches, the signal amplitude is expected to be small in comparison to the noise and its frequency could be changing unpredictably over time.
A Fourier transform of the entire dataset splits the signal power between multiple frequency bins and could result in the signal being undetectable under the noise (see Fig.~\ref{fig:viterbi_comparison_webcam_spectrum}). 
The Viterbi algorithm's strength lies in its ability to track such signals through the data.
In the following section, we present the results of using the Viterbi algorithm with the table-top interferometer data. 
% - just simiplifed this discussion a little and added a reference to the CW searches
%The key to the Viterbi algorithm is that we can use the signal's correlation in time to improve the signal-to-noise ratio of the frequency of the wandering signal at a particular time. Since the noise is assumed to not be correlated in time, if the detection statistic at a given frequency is high at a particular time and at neighboring frequencies at neighboring times then the likelihood that the given frequency at that particular time is from the wandering signal increases. By iterating this process through the grid for a suitable choice of the window of neighboring frequencies and times, the Viterbi algorithm is well-suited to finding wandering signals.
